Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-02-15 paper/letter.tex
editando

The first generation of galaxies point to our cosmic corigins. Their
initial conditions and characteristiscs are the first page on the
evolution of galaxies such as the milky way that evolve to the present
time. In the standard big bang cosmology the only elemets that were
created in the nucleosynthesis where Hydrogen, Helium and
Lithium. Heavier elemets must hve been created in the internal
processes of stellar evolution. Therefore the first generation of
galaxies shoudl be metal free and rich in hydrogen. We haven detected
these galaxies yet. The best obsrevational proxy we have for them are
dwarf star forming galaxies with a low metallicity content. Compact
dwarf galaxies (CDGs) are the best observational example.

How can we develop a strategy to look for the first galaxies How can we recognize a first galaxy when we see one? 
Almost fifty years ago \cite{PartridgePeebles} realized that young
galaxies could be detected through a strong Lyman-$\alpha$ line
emission.  This theoretical prediction was only confirmed XX year
later.  Currently, Lyman Alpha Emitting galaxies are commonly measure
in surveys. 
The presence of this line provides confirmation of the distance of a
galaxy and also provides clues about the stellar population powering
the emission.

Lyman alpha emission, as suggested in the first paragraph, is not
exclusive of high redshift galaxies. Any galaxy with low metallicity
and dust content and ongoing star formation should present this line.  This has motivated the creation of
local Universe surveys to perform a detailed study of the Lyman-alpha
emission \cite{LARS}. A nearby simple permits the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

The Lyman-$\alpha$ line is resonant. A yman-$\alpha$ photon
will follow a difussion-like process before escaping the galaxy or
being absorbed by dust. This makes it hard to interpret Lyman-alpha
line observations as the reulting profile is sensitive to the
dynamical, chemical and thermal conditions in the interstellar
medium. State-of-the-art models use montecarlo radiative transfer
simulations to interpret lya line profiles. Only in very few symmetric
cases analytical solutions are available.

@@

The Ly$\alpha$ emission line was first observed in the \tol
galaxy by \cite{Thuan97}. It has two main important features which 
make this a very uncommun LAE. First it has symmetric profile 
without any systemic displacement from the expected
line-center. Second it shows a triple peak, while the most common
observational feature corresponds either to a single or asymetric
double peak. 
This profile has escaped a clear theoretical understading based on
models that consider galaxies as spherical gas distributions with an
outflow. 
We have attempted an interpretation of the lya spectrum in \tol\ wihas
a central feature the bulk gas rotation without any outflow. This
model provides a complete explanation of \tol's puzzling feature.

>>>>

The first generation of galaxies trace our cosmic origins. 
Their characteristiscs are the first step on the evolution of galaxies
such as the Milky Way.
In the standard Big Bang cosmology the only elements that were
created in the nucleosynthesis process were Hydrogen, Helium and
Lithium. 
Heavier elemets must have been created in stellar evolution process. 
Therefore the first generation of
galaxies should be metal free and rich in Hydrogen. 
This kind of primordial galaxies have not been detected yet. 
However, dwarf star forming galaxies with a low metallicity content
are close template to understand the early galaxy evolution process.

Almost fifty years ago \cite{PartridgePeebles} it was realized that
young galaxies could be detected through a strong Lyman-$\alpha$ line
emission.  
This theoretical prediction was only confirmed thirty year later,  not
on first generation galaxies but on distant relatively young systems. 
Currently, Lyman Alpha Emitting (LAE) galaxies are commonly targetted
in surveys. 
The presence of the Ly-$\alpha$ emission line provides confirmation of
the distance of a galaxy and also provides clues about the stellar
population powering the emission.

Ly-$\alpha$ emission is not exclusive of distant galaxies. 
Any galaxy with low dust content and ongoing star formation has the
right conditions to show this line.  
There are, for instance,  local Universe surveys that target
Ly-$\alpha$ emission in nearby dwarf star forming galaxies. 
\cite{LARS}. 
A nearby LAE sample allow for the study of other indicators that
might be more difficult to obtain for distant galaxies such as
morphology, dust attenuation, neutral hydrogen contents and ionization
state.

However, the physical interpratation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon followw a difussion-like process before escaping
the galaxy or being absorbed by dust. 
The reulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
Only in very few cases of high symmetry there are analytical solutions
available. 
Interpretation of observations require state-of-the-art models use
montecarlo radiative transfer simulations.

@@

\tol is a compact star forming dwarf galaxies (CDGs) that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two features that had
so far evaded a natural physical interpratation.
First, it has symmetric profile  without any systemic displacement
from the expected line-center. 
Second, it shows a triple peak, while the most common observational
feature corresponds either to a single or asymetric double peak. 
We have attempted an interpretation of the Ly-$\alpha$ line in
\tol\ by including the effect of gas bulk rotation.
We find that this model provides a satisfactory explanation of \tol's
puzzling features.
